---
title: "cm008 Exercises"
output: github_document

---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))

```

## `mutate()`

Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}
<<<<<<< HEAD
gapminder %>% 
  mutate(gdp = gdpPercap*pop,
         gdpBill=round(gdp/10^9,2))
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```


Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
<<<<<<< HEAD
gapminder %>% 
  transmute(gdp = gdpPercap*pop,
         gdpBill=round(gdp/10^9,2))
```

The `if_else` function is useful for changing certain elements in a data frame.


```{r}
#example of if_else
if_else(letters == "h", "H", letters)

```


=======

```

The `if_else` function is useful for changing certain elements in a data frame.
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}
<<<<<<< HEAD
gapminder %>% 
  mutate(lifeExp = if_else(country == "Canada"&year==1952,70, lifeExp)) %>% 
  filter(country == "Canada")
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```


Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
<<<<<<< HEAD
gapminder %>% 
  mutate(cc= paste(country, continent, sep=", "))
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

These functions we've seen are called __vectorized functions__. 

## `summarize()` and `group_by()`

Use `summarize()` to compute the mean and median life expectancy using all entries:

```{r}
<<<<<<< HEAD
gapminder %>% 
  summarize(mu=mean(lifeExp),
            md=median(lifeExp))
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

Do the same thing, but try:

1. grouping by country
2. grouping by continent and country

```{r}
<<<<<<< HEAD
gapminder %>% 
  group_by(country) %>% 
  summarize(mu= mean(lifeExp),
            md=median(lifeExp))


```

```{r}
gapminder %>% 
  group_by(country, continent) %>% 
  summarize(mu= mean(lifeExp),
            md=median(lifeExp))
```
=======

```

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
- Notice the columns that are kept. 
- Notice the grouping listed above the tibble, especially without a call after grouping.
- Notice the peeling of groups for each summarize. 

Question: What if I wanted to keep the other numeric columns (gdpPercap, pop)? Can I? Would this even make sense?
<<<<<<< HEAD
*Difference between dimensions*
=======
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

For each continent: What is the smallest country-wide median GDP per capita?

```{r}
<<<<<<< HEAD
gapminder %>% 
  group_by(continent,country) %>% 
  summarize(md=median(gdpPercap)) %>% 
  summarize(min=min(md))
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

Note that ggplot2's grouping is different from dplyr's! Try making a spaghetti plot of lifeExp over time for each coutry, by piping in a grouped data frame -- it won't work:

```{r}
<<<<<<< HEAD
gapminder %>% 
  group_by(country) %>% 
  ggplot(aes(year,lifeExp))+ #ggplot dont recognize `group_by`
  geom_line()
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```


Your turn! For each continent, what is the median GDP per capita of countries with high (>60) life expectancy vs countries with low (<=60)? Sort this data frame by median GDP per capita.

```{r}
<<<<<<< HEAD
gapminder %>% 
  mutate(hvsl=if_else(lifeExp>60, "high","low")) %>% 
  #group_by(continent,lifeExp>60) %>% 
  group_by(continent,hvsl) %>% 
  summarize(md = median(gdpPercap)) %>% #2 group for each continent
  arrange(md)


gapminder %>%
  mutate(age=if_else(lifeExp >60,"high","low")) %>% 
  group_by(continent,age) %>% 
  summarize(md =median(gdpPercap))
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

There are special functions to summarize by. Let's see some of them:

- `n()`: Number of rows in the group.
- `n_distinct()`

Convenience functions:

- `tally()` (= `summarize(n = n())`)
- `count(...)` (= `group_by(...) %>% tally()`)

n_distinct: How many years of record does each country have?

```{r}
<<<<<<< HEAD
gapminder %>% 
  group_by(continent) %>% 
  #summarize(num= n())
  tally()

gapminder %>% 
  count(continent)
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

count

## Function types

Let's stop to identify some theory of function types, and the `dplyr` approach to them.

| Function type | Explanation | Examples | In `dplyr` |
|------|-----|----|----|
| Vectorized functions | These take a vector, and operate on each component independently to return a vector of the same length. In other words, they work element-wise. | `cos`, `sin`, `log`, `exp`, `round` | `mutate` |
| Aggregate functions | These take a vector, and return a vector of length 1 | `mean`, `sd`, `length`, `typeof` | `summarize`, esp with `group_by`. |
| Window Functions | these take a vector, and return a vector of the same length that depends on the vector as a whole. | `lag`, `rank`, `cumsum` | `mutate`, esp with `group_by` |

For any generic output, we can use dplyr's `do()` function -- but that's a topic for STAT 547. 
<<<<<<< HEAD
```{r}
rank(c(4,5,3,2,5,5,2,1,6,7)) #average rank to a tie
```
=======
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

## Grouped `mutate()`

Calculate the growth in population since the first year on record _for each country_. `first()` is useful.

```{r}
<<<<<<< HEAD
gapminder %>% 
  group_by(country) %>% 
  mutate(growth = pop - first(pop))


gapminder %>% 
  group_by(country) %>% 
  mutate(change = pop - lag(pop))

=======
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

```

Notice that `dplyr` has retained the original grouping.

How about growth compared to `1972`?

```{r}

<<<<<<< HEAD
gapminder %>% 
  group_by(country) %>% 
  mutate(change = pop - pop[year==1972])

=======
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

Make a new variable `pop_last_time`, as the "lag-1" population -- that is, the population from the previous entry of that country. Use the `lag` function.

```{r}

```

Similar: `lead` function.

Notice the NA's. 

## Putting it all together

Your turn: Use what we learned to answer the following questions.

1. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent.

```{r}
<<<<<<< HEAD
gapminder %>% 
  group_by(continent,country) %>% 
  mutate(gain= lifeExp-lag(lifeExp)) %>% #all 5 year apart
  filter(!is.na(gain)) %>% #not nes
  summarize( min =min(gain)) %>% 
  summarize(min=min(min)) %>% 
  arrange(min)
           
=======

>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11
```

2. Compute the relative gdp (NOT per capita!) of each country compared to Canada (= GDP of a country / GDP of Canada).

```{r}
<<<<<<< HEAD
cda = 
  gapminder %>% 
  mutate(gdp = gdpPercap * pop) %>% 
  filter(country == "Canada") %>% 
  '[['("gdp")

my_df=
  gapminder %>% 
  mutate(cda_gdp=rep(cda,nlevels(country)),gdp=gdpPercap*pop,
         rel_gdp=gdp/cda_gdp)  
  
my_df %>% filter(country=="Canada")

summary(my_df$rel_gdp)
=======
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

```

Sanity check: are Canada's numbers = 1? What is the spread of numbers like (should be small)?

## Summary of major one-table functions

- `select()`
- `filter()`
- `arrange()`
<<<<<<< HEAD
- `mutate()` 
- `summarize()` 
=======
- `mutate()`
- `summarize()`
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

Together with `group_by()` and "twists" of the above. 


## Practice Exercises

Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.

1. Convert the population to a number in billions.

2. Compute the change in population from 1962 to 1972 for each country.
<<<<<<< HEAD
```{r}
gapminder %>% 
  filter(pop>=1962 & pop <=1972) %>% 
  group_by(country) %>% 
  mutate(change = pop -lag(pop,2))
```
=======
>>>>>>> 2c8be30b8d57fe3efdb168406825070edd459b11

3. Rank the continents by GDP per capita. You should have two columns: one with continent, and another with the ranking (1 through 5). __Hint__: use the `rank()` or `min_rank()` function.
