---
title: "Data Frame Exploration"
output: 
    html_document:
        theme: cerulean
        toc: true
        keep_md: true
---

## LaTeX Equations in R Markdown

Here is an equation:

$$\alpha = 5 + 2$$

Inline equation: $\alpha = 5 + 2$. 

## Data frame exploration

### Setting up

First, load the `gapminder` R package. If you don't have it installed, run `install.packages("gapminder")` in the console first.

install.packages("gapminder")


```{r}
install.packages("gapminder")
library(gapminder)
```

### What is a data frame?

It's tabular data. To get a sense of this, go ahead and print out the `gapminder` object (you might want to do this in the console!).
```{r}
gapminder

```
### Exploration of data frames

Let's explore `gapminder` with functions like `head`, `ncol`, `str`, `summary`.
```{r}
head(gapminder) #top 6 rows 

```
```{r}

str(gapminder) # dataframe about this data
ncol(gapminder) #number of cols
summary(gapminder)# statistical summary 
```


### Extracting columns/"variables"

Let's extract a column with `$`. Maybe compute its variance.
```{r}

gapminder$country 

```


```{r}
gapminder[1:4] #extract cols

gapminder[1:4,]#extract rows

```